---
title: "STAC67 A1"
output:
  pdf_document: default
  html_document: default
date: "2026-01-14"
---

# Q1

## (a)

```{r results='hold'}
csvFile = "OlympicMarathon2016.csv"
df = read.csv(csvFile)
mu = mean(df[["Minutes"]])
sigma2 = var(df[["Minutes"]])
sprintf("Minutes has mean %f and variance %f", mu, sigma2)
```

```{r}
hist(df[["Minutes"]], xlab="Minutes", main="histogram of the Minutes")
```

## (b)

```{r results='hold'}
set.seed(1008494620)
population = df[["Minutes"]]
sampleCnt = 10000
n = 30
yBar = numeric(sampleCnt)
stats = numeric(sampleCnt)
for (i in 1:sampleCnt) {
  y = sample(population, size=n, replace=TRUE)
  yBar[i] = mean(y)
  s2 = var(y)
  stats[i] = (n-1) * s2 / sigma2
}
```

### (b)(i)

```{r results='hold'}
mean_yBar = mean(yBar)
var_yBar = var(yBar)
mean_stats = mean(stats)
var_stats = var(stats)

percentiles = c(0.025, 0.25, 0.5, 0.75, 0.975)
percentiles_pop = mapply(
  function(p) {
    quantile(df[["Minutes"]], p)
  },
  percentiles
)
percentiles_yBar = mapply(
  function(p) {
    quantile(yBar, p)
  },
  percentiles
)
percentiles_stats = mapply(
  function(p) {
    quantile(stats, p)
  },
  percentiles
)

cat(
  sprintf("theoretical values\n"),
  sprintf("==================\n"),
  sprintf("mean = %f, var = %f\n", mu, sigma2),
  sprintf("percentiles:\n"),
  percentiles_pop
)

cat(
  sprintf("\n\nresult of sample means\n"),
  sprintf("======================\n"),
  sprintf("mean = %f, var = %f\n", mean_yBar, var_yBar),
  sprintf("percentiles:\n"),
  percentiles_yBar
)

cat(
  sprintf("\n\nresult of scaled sample var\n"),
  sprintf("===========================\n"),
  sprintf("mean = %f, var = %f\n", mean_stats, var_stats),
  sprintf("percentiles:\n"),
  percentiles_stats
)
```

### (b) (ii)

```{r results='hold'}
hist(yBar,
  probability = TRUE,
  main="sampling distribution of yBar",
  xlab="yBar",
)
x1 = seq(min(yBar), max(yBar), length.out = 500)
lines(x1, dnorm(x1, mean=mu, sd=sqrt(sigma2/n)), lwd=2)

hist(stats,
  probability=TRUE,
  main="sampling distribution of scaled sample variance",
  xlab=expression(bar(y)),
)
x2 = seq(0, max(stats), length.out = 500)
lines(x2, dchisq(x2, df=n - 1), lwd=2)
```

# Q7

## (a)

```{r results='hold'}
library(ggplot2)
csvFile = "kidiq.csv"
df = read.csv(csvFile)
ggplot(
  df, 
  aes(
    x = mom.iq, 
    y = kid.score,
  )
) + geom_point() + geom_smooth(formula = y~x, method = "lm")

```

## (b)

```{r results='hold'}
y = df[["kid.score"]]
x = df[["mom.iq"]]
fit = lm(y~x, data = df)
beta0 = fit$coefficients[["(Intercept)"]]
beta1 = fit$coefficients[["x"]]
cat(sprintf("beta0 = %f, beta1 = %f", beta0, beta1))
```

## (c)

```{r results='hold'}
x_bar = mean(x)
y_bar = mean(y)

SS_yy = sum((y - y_bar)^2)
SS_xy = sum((y - y_bar) * (x - x_bar))
SS_xx = sum((x - x_bar)^2)

beta1_hat = SS_xy / SS_xx
beta0_hat = y_bar - beta1_hat * x_bar

cat(
  sprintf("R built-in function\n"),
  sprintf("==================\n"),
  sprintf("intercept = %f, slope = %f\n", beta0, beta1),
  
  sprintf("\nComputed result\n"),
  sprintf("===============\n"),
  sprintf("beta0_hat = %f, beta1_hat = %f", beta0_hat, beta1_hat)
)
```

**Interpretation:**

-   $\hat{\beta_0} = 25.799778$ implies, in theoretical speaking, the children's test scores is 25.799778 when their mother's IQ scores is 0.

-   $\hat{\beta_1} = 0.609975$ implies that children's test scores are positively correlated with their mother's IQ scores – every increment in the one unit of their mother's IQ scores will result in a raise of 0.6009975 on children's test scores.

## (c)

```{r result='hold'}
alpha = 0.05
n = length(x)

sse = SS_yy - beta1_hat * SS_xy
ss = sse / (n - 2)
se_beta1 = sqrt(ss / SS_xx)
se_beta0 = sqrt((1/n + x_bar^2/SS_xx) * ss)

# two-sided CI
tcrit = qt(1 - alpha/2, df = n - 2)

beta0_ci_upper = beta0_hat + tcrit * se_beta0
beta0_ci_lower = beta0_hat - tcrit * se_beta0

beta1_ci_upper = beta1_hat + tcrit * se_beta1
beta1_ci_lower = beta1_hat - tcrit * se_beta1

cat(
  sprintf("handcrafted CIs\n"),
  sprintf("===============\n"),
  sprintf("            2.5%%      97.5%%\n"),
  sprintf("beta0: %f, %f\n", beta0_ci_lower, beta0_ci_upper),
  sprintf("beta1: %f, %f\n", beta1_ci_lower, beta1_ci_upper),
  
  sprintf("\n\nbuilt-in function\n"),
  sprintf("=================\n")
)
confint(fit)
```

**Interpretation:**

-   For $\beta_1$: On every increment of the mother's IQ scores, the average response of the children's test scores is increased by between 0.4949534 and 0.7249957 on average. However, in the real world practice, having a IQ score of 0 is meaningless, the interpretation is therefore a theoretical baseline under the scene of statistics study, instead of an important quantity.

-   For $\beta_0$: When the mother's IQ scores is zero, the children's test scores are expected to between 14.1692789 and 37.4302768.

-   For the constructed confidence intervals: When the procedure of random sampling is performed under the same population, 95% of the constructed intervals are going to contains the true parameter $\beta_0$ and $\beta_1$.

## (d)

```{r results='hold'}
t = beta1_hat / se_beta1
tcrit = qt(0.95, df = n-2)
cat(sprintf("test stat: %f, the one-sided critical t is given by: %f\n", t, tcrit))
if (t > tcrit) {
  cat("We reject H0. Evidence suggests a positive association.\n")
} else {
  cat("We fail to reject H0. Insufficient evidence of a positive association at this alpha.\n")
}
p_val <- 1 - pt(t, df = n-2)
cat(sprintf("p-value is given by: %f\n", p_val))
```

## (e)

```{r results='hold'}
tt = 0.1 / se_beta1
p = 2 * (1 - pt(tt, df = n-2))
sprintf("the probability is: %f", p)
```
